We next address the problem of \emph{unsupervised} learning. 
In this setting, we are not given any labeled data; all we get to see is a set of natural language sentences.  
The underlying question is: 
\begin{quote}
Can we learn something from raw text?
\end{quote}
This task is challenging, since the process by which linguistic structures are generated is not always clear; 
and even when it is, it is typically too complex to be
formally expressed. Nevertheless, unsupervised learning has been applied to a
wide range of NLP tasks, such as: 
\pos\ Induction  \citep{schutze1995distributional,merialdo1994tet,clark03combining},
Dependency Grammar Induction \citep{klein2004acl,smith2006annealing}, Constituency Grammar Induction \citep{klein2004acl}, Statistical Word Alignments 
\citep{brown94mathematic} and Anaphora Resolution \citep{charniak2009works}, just to name a few. 

Different motivations have pushed research in this area. From both a linguistic and cognitive point of view, 
unsupervised learning is useful as a tool to study language acquisition. 
From a machine learning point of view, unsupervised learning is a fertile ground for testing new learning methods, 
where significant improvements can yet be made. 
From a more pragmatic perspective, unsupervised learning is required
since annotated corpora is a scarce resource for different reasons. Independently of the reason, unsupervised learning is an increasingly active field of research.

A first problem with unsupervised learning, since we don't observe any labeled data (i.e., 
the training set is now $\mathcal{D}_{U} = \{x^1,\ldots, x^M\}$), 
is that most of the methods studied so far (e.g., Perceptron, MIRA, SVMs) cannot be used since we cannot compare 
the true output with the predicted output. 
Note also that a direct minimization of the \emph{complete negative log-likelihood} of the data, $\log P_{\theta}(X^1=x^1,\ldots,X^M=x^m)$, 
is very challenging, since it would require marginalizing out (\emph{i.e.}, summing over) all possible hidden variables:
\begin{equation}
 \log P_{\theta}(X^1=x^1,\ldots,X^M=x^m) =  \sum_{m=1}^M \log \sum_{y \in \Lambda} P_{\theta} (X=x^m,Y=y).
\end{equation}
Note also that the objective above is \emph{non-convex} even for a linear model: hence, it may have local minima, which makes optimization much 
more difficult.%
%\footnote{Another observation is that normally we are restricted to generative models, 
%with some remarkable exceptions~\citep{smith2005acl}, since the objective of discriminative models when no labels are observed are 
%meaningless ($\sum_{y^m } P(y^m |x^m) = 1$); this rules out, for instance, Maximum Entropy classifiers.}

The most common optimization method in the presence of hidden (latent) variables is the Expectation Maximization (EM) algorithm, described in the next section. Note that this algorithm is a generic optimization routine that does not depend on a particular model. Later, in Section \ref{posi} we will apply the EM algorithm to the task of part-of-speech induction, where one is given raw text and a number of clusters and the task is to cluster words that behave similarly in a grammatical sense. 

\subsection{\label{em}Expectation Maximization Algorithm}
\input{pages/sequences/em.tex}



\subsection{\label{posi}Part of Speech Induction}
\input{pages/sequences/pos-induction.tex}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../../guide"
%%% End: 
